---
title: "COVARIATES (update: sep2016.dta)"
author: "Andree Valle Campos"
date: '`r Sys.Date()`'
output:
  #html_notebook:
  html_document:
    fig_caption: yes
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: yes
    code_folding: "hide"
    #df_print: paged
bibliography: malaria.bib
#csl: american-medical-association.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, #fig.path = "01-",
                      warning = FALSE)
knitr::opts_knit$set(root.dir = '../.')

options(width = 110) # expand limits of CONSOLE output

require(Hmisc)
mu <- markupSpecs$html   # markupSpecs is in Hmisc
# hidden command (<code>r mu$widescreen()</code>) to use an entire wide screen. # mu$widescreen()
```
`r mu$widescreen()`

# Aim

Retrieve relevant variables from the *Severe Vivax Malaria study* 
to use them as covariates for the subset of samples provided to the 
*Protein microarray Ab profile study*.

# Done

- **new** variables: 
    + `glas_11`: Glasgow score lower than 11 (and different from 0)
    + `sev_WHO_num`: Number of Severe Malaria WHO criterias of the patient
    + `sev_WHO`: Absence or presence of Severe Malaria WHO criterias of the patient
- `glas_11` [here](#tidy-data):
    + Glasgow < 11 & !=0
    + **differs** from `cns_malaria`
- `sev_WHO` [here](#filter-who-criteria):
    + **includes:** 
        + `glas_11`, `sdra`, `lung_injury`, `ictericia_corr`, `hiperbil`, `shock`, 
        `bleeding`, `glic_normal`, `hiperparasitemia`, `anemia_severa`.
    + **Not** included `cns_malaria`
    + **differs** from `cas_con_admin2`
- Inferential statistics [here](#inferential-statistics) (exploratory):
    + given a stratification, the following variables were significatly different:
        + **`sev_WHO`**: proportion of `sex` and `malaria_pasado`
        + **`malaria_pasado`**: mean `age` and proportion of `sev_WHO`
- Ab profile subset of **60 samples** [here](#ab-profile-subset) :
    + frequency tables with respect to severity, previous episodes and combined.

# To do

- **review** new variables `glas_11` and `sev_WHO`
- `sev_WHO`:
    + **include** as **renal impairment / failure criteria** `hemoglobinuria_corr` or `oliguria_corr`?
    + **meaning** of `cas_con_admin2`?
- **change** `NA` to `0` among severe [covariates](#proportions-per-criteria)?

# Notes

- From this dataset:
    + Since all the cases were reported in a Hospital, can all patients be considered symptomatic?
    + Although the proportion of `malaria_pasado` significatly differs w.r.t. `sev_WHO`,
    *previous episodes of malaria* may not be directly associated with `sev_WHO`.
    + Continuous Exposure (rather than previous reported cases of malaria) could be 
    a *conditional* factor to achieve immunity (asympomatic),
    but it may not be *determinant* to protect against severity.

- `malaria_pasado` limitations:
    + is self-reported.
    + is not exposure.
    + could be validated/improved with MSP1 ELISA quantification.
    + more relevant would be to take into account *time since last exposure (or malaria event)* 
    in a follow-up longitudinal study.

# Analysis

## Dependencies
```{r, message=FALSE}
##essential
library("tidyverse")  # load dplyr, tidyr, tibble, readr, purr, ggplot2
library("haven")      # import STATA .dta files
library("Hmisc")      # Data dictionary + descriptive / stratified statistics
##extra
library("forcats")    # factor vectors
library("stringr")    # for strings
library("knitr")      # To display nice tables
library("Rmisc")      # multiplot ggplots
library("viridis")    # color visualization

```

## Severe dataset

### Input

- Read `severe_malaria_vivax SEP2016.dta` file: 

```{r}
sevdb <- haven::read_dta("data-raw/severe_malaria_vivax SEP2016.dta") #%>% haven::as_factor() #ERROR
```

### Data dictionary

- Including the **labels** from the `.dta` file:

```{r}
#Hmisc::contents(sevdb)
html(Hmisc::contents(sevdb), maxlevels=10, levelType='table')
#sevdb %>% select(contains("fiebre")) %>% group_by(fiebre_hoy_corr,fiebre_us_corr) %>% dplyr::count()
```

### Tidy data

- **create** `glas_11` 
    + Glasgow < 11 & !=0
    + **differs** from `cns_malaria`

- sobre `fiebre`: preferencia en `_hoy_corr`

- hay variable `coinfections`?

```{r}
# NO creatinina -> ???
# NO parasitemia -> hiperparasitemia
sevdb_1 <- sevdb %>% 
  select(codigo, edad, sexo, cas_con_admin2,
         eritro_parasit, admin_uci_corr:temperatura, 
         ends_with("_corr"), -ends_with("_us_corr"), ends_with("_hoy_corr"), 
         presenta_comorb, starts_with("enf_"), diabetes,
         ends_with("_riesgo"), starts_with("malaria_"),
         -glas_ojos, -glas_verbal, -glas_motor, -glas_total, -num_convulsion, 
         -bilirrubina, -cont_plaquetas,
         -starts_with("droga"), -starts_with("tx_")
         ) %>% 
  mutate(codigo= str_replace(codigo,"LIM0","LIM"),
         
         glas_11= if_else(glas_corr<11 & glas_corr>0,1,0),
         
         shock= str_replace(shock,"NaN|99",replacement = NA_character_),
         sdra= str_replace(sdra,"NaN|99",replacement = NA_character_),
         hiperbil= str_replace(hiperbil,"NaN|99",replacement = NA_character_),
         malaria_pasado_num=  str_replace(malaria_pasado_num,"98|99",replacement = NA_character_),
         fiebre_hoy_corr=  str_replace(fiebre_hoy_corr,"99",replacement = NA_character_),
         cas_con_admin2=  str_replace(cas_con_admin2,"99",replacement = NA_character_)
         ) %>% 
  mutate(shock= as.double(shock),
         sdra= as.double(sdra),
         hiperbil= as.double(hiperbil),
         malaria_pasado_num= as.numeric(malaria_pasado_num),
         malaria_pasado= as.factor(malaria_pasado),
         fiebre_hoy_corr= as.numeric(fiebre_hoy_corr),
         cas_con_admin2= as.numeric(cas_con_admin2)
         ) %>% 
  mutate(malaria_pasado= fct_recode(malaria_pasado, "con-episodio-previo"="1", "sin-episodio-previo"="0"))

#glimpse(sevdb_1)
#sevdb_1 %>% dplyr::count(malaria_pasado)
#Hmisc::contents(sevdb_1)

sevdb_1 %>% group_by(glas_11, glas_corr, cns_malaria) %>% dplyr::count()
```

#### Check updates

- **CNS malaria** = postracion, coma, convulcion, confusion
- **LUNG_INJURY** = at least 2 criteria of lung related-injuries (disnea, taquipnea, infiltracion, edema)
- **BLEEDING** = at least one criteria of bleeding (gingivorragia_corr, epistaxis_corr, sang_intest_corr)
- **HIPOGLICEMIA** = glic_normal = hipo_glic1 + hipo_glic2
- DIFFERENT VARIABLES
    + ictericia!=hiperbili
- **COMORBILITIES** = diabetes, enf_cardiac, enf_pulm, enf_renal, enf_snc, enf_otra

```{r}
# CNS malaria = postracion, coma, convulcion, confusion
sevdb_1 %>% group_by(cns_malaria,postracion_corr, coma_corr,convulsion_corr,confusion_corr) %>% dplyr::count()

# SDRA != LUNG_INJURY = at least 2 criteria of lung related-injuries (disnea, taquipnea, infiltracion, edema)
sevdb_1 %>% group_by(lung_injury,disnea_corr,taquipnea_corr,infiltracion_corr,edema_pulm_corr) %>%  dplyr::count()

# BLEEDING = at least one criteria of bleeding (gingivorragia_corr, epistaxis_corr, sang_intest_corr)
sevdb_1 %>% group_by(bleeding,gingivorragia_corr,epistaxis_corr,sang_intest_corr) %>%  dplyr::count()

# HIPOGLICEMIA
sevdb_1 %>% group_by(glic_normal,hipo_glic1,hipo_glic2) %>%  dplyr::count()   # glic_normal = hipo_glic1 + hipo_glic2

# DIFFERENT VARIABLES
sevdb_1 %>% group_by(ictericia_corr,hiperbil) %>%  dplyr::count()             # ictericia!=hiperbili
#sevdb_1 %>% group_by(admin_uci_corr,dias_uci) %>% dplyr::count()

# COMORBIDITIES
sevdb_1 %>% group_by(presenta_comorb, diabetes, enf_cardiac, enf_pulm, enf_renal, enf_snc, enf_otra) %>%  dplyr::count()
```

### Filter WHO criteria

Según @WHOsevere2014:

- **create** `sev_WHO`
    + **includes:** 
        + `glas_11`, `sdra`, `lung_injury`, `ictericia_corr`, `hiperbil`, `shock`, `bleeding`, `glic_normal`, `hiperparasitemia`, `anemia_severa`.
    + **Not** included `cns_malaria`
    + **differs** from `cas_con_admin2`

- deberia incluir como **renal impairment / failure**:
    + `hemoglobinuria_corr`
    + `oliguria_corr`

- referenciales:
    + `eritro_parasit`
    + `admin_uci_corr`, `dias_uci`
    + `fiebre_hoy_corr`
    + `presenta_comorb`
    + `malaria_pasado`, `malaria_pasado_num`

```{r}
sevdb_2 <- sevdb_1 %>% 
  select(codigo:eritro_parasit,
         contains("uci"), 
         fiebre_hoy_corr,
         cns_malaria,
         
         glas_11,
         sdra,
         lung_injury,
         ictericia_corr, hiperbil,
         shock,
         bleeding,
         glic_normal,
         hiperparasitemia,
         anemia_severa, #hemoglobinuria_corr,
         
         presenta_comorb,
         malaria_pasado, malaria_pasado_num
         ) %>% 
  mutate(sev_WHO_num= rowSums(.[10:19],na.rm = T), # OJO: de glas_11 -> anemia_severa
         sev_WHO= if_else(sev_WHO_num>0,"severo","no-severo")
         ) %>%
  mutate(sev_WHO= as.factor(sev_WHO))

html(Hmisc::contents(sevdb_2), maxlevels=10, levelType='table')
#glimpse(sevdb_2)

sevdb_2 %>% group_by(malaria_pasado, malaria_pasado_num) %>% dplyr::count()
sevdb_2 %>% dplyr::count(malaria_pasado)
sevdb_2 %>% dplyr::count(sev_WHO)
sevdb_2 %>% group_by(cas_con_admin2, sev_WHO) %>% dplyr::count()
sevdb_2 %>% group_by(malaria_pasado, sev_WHO) %>% dplyr::count()

#sevdb_2 %>% group_by(cns_malaria,glas_11) %>% dplyr::count()

```

#### proportions per criteria

```{r}
s1 <- Hmisc::summaryM(glas_11 +
                        sdra +
                        lung_injury +
                        ictericia_corr + hiperbil +
                        shock +
                        bleeding +
                        glic_normal +
                        anemia_severa +
                        hiperparasitemia + #parasitemia + 
                        sev_WHO_num +
                        fiebre_hoy_corr
                      ~ sev_WHO
                      ,
               data=sevdb_2,
               overall=FALSE, test=FALSE)

latex(s1, caption='Laboratory data',
      exclude1=TRUE, npct='both', 
      digits=3,
      prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
      middle.bold=TRUE, long = TRUE,
      #legend.bottom = TRUE, #insert.bottom = TRUE, 
      what="%", html = TRUE
      ) #change here for LaTeX PDF
```

#### combination of criterias

```{r}
# SUM SEVERITY CRITERIAS
sevdb_o <- sevdb_2 %>% 
  dplyr::rename(glas=glas_11,
         #sdra,
         lung=lung_injury,
         icter=ictericia_corr,
         bili=hiperbil,
         #shock,
         bleed=bleeding,
         glic=glic_normal,
         paras=hiperparasitemia,
         anemia=anemia_severa#, 
         #hemo=hemoglobinuria_corr
         ) 
sevdb_o %>% 
  group_by(#cns_malaria,
           glas,
           sdra,
           lung,
           icter,
           bili,
           shock,
           bleed,
           glic,
           paras,
           anemia#, 
           #hemo
           ) %>% 
  dplyr::count() %>% 
  dplyr::arrange(desc(n))
```

```{r}
sevdb_o %>% 
  select(#cns_malaria,
           glas,
           sdra,
           lung,
           icter,
           bili,
           shock,
           bleed,
           glic,
           paras,
           anemia#, 
           #hemo
           ) %>% 
  summarise_all(sum,na.rm=T) #%>% t()
```

```{r, fig.height=4, fig.width=6, fig.align='center', eval=FALSE}
## symptoms x exposure

#plot(sevdb_2$malaria_pasado_num,sevdb_2$sev_WHO_num)
sevdb_2 %>% 
  group_by(malaria_pasado_num, sev_WHO_num) %>%
  dplyr::summarise(n=n()) %>%
  mutate(prop= n/sum(n)) %>%
  ggplot(aes(malaria_pasado_num, sev_WHO_num, colour=n)) + 
  geom_point() + 
  scale_color_viridis() +
  labs(title="sevdb_2")
```

### Inferential statistics

- Joint `parasitemia` from `28Nov2014_Ed.dta` **[WARNING]**

```{r}
sevdb_x <- haven::read_dta("data-raw/malaria severa 28Nov2014_Ed.dta") %>% 
  select(codigo, parasitemia)%>% 
  mutate(codigo= str_replace(codigo,"LIM0","LIM"))

sevdb_2x <- sevdb_2 %>% 
  haven::as_factor() %>% 
  dplyr::inner_join(sevdb_x,by="codigo")
```

***

**wrt severity**

- significant differences with respect to 
    + proportion of previous episodes of malaria,
    + proportion of women patients towards severe cases

```{r}
s1 <- Hmisc::summaryM(edad + sexo +
                        malaria_pasado + #malaria_pasado_num + 
                        #sev_WHO_num #+ sev_WHO
                        fiebre_hoy_corr + presenta_comorb +
                        parasitemia #+
                      ~ sev_WHO #+ 
                      #malaria_pasado
                      ,
               data=sevdb_2x,
               overall=FALSE, test=TRUE)

latex(s1, caption='Clinical data',
      exclude1=TRUE, npct='both', 
      digits=3,
      prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
      middle.bold=TRUE, long = TRUE,
      #legend.bottom = TRUE, #insert.bottom = TRUE, 
      what="%", html = TRUE
      ) #change here for LaTeX PDF

```

***

**wrt previouse episodes of malaria**

- significant differences with respect to 
    + mean age,
    + proportion of severe patients

```{r}
s1 <- Hmisc::summaryM(edad + sexo + 
                        #malaria_pasado_num + #malaria_pasado +
                        sev_WHO + sev_WHO_num +
                        fiebre_hoy_corr + presenta_comorb +
                        parasitemia #+
                      ~ #sev_WHO #+ 
                      malaria_pasado
                      ,
               data=sevdb_2x,
               overall=FALSE, test=TRUE)

latex(s1, caption='Clinical data',
      exclude1=TRUE, npct='both', 
      digits=3,
      prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
      middle.bold=TRUE, long = TRUE,
      #legend.bottom = TRUE, #insert.bottom = TRUE, 
      what="%", html = TRUE
      ) #change here for LaTeX PDF
```

***

**wrt severity by previous episodes of malaria**

- no significant differences with respect to any variable

```{r}
s1 <- Hmisc::summaryM(edad + sexo +
                        #malaria_pasado_num + #malaria_pasado +
                        #sev_WHO_num #+ sev_WHO
                        fiebre_hoy_corr + presenta_comorb +
                        parasitemia #+
                      ~ sev_WHO + 
                      malaria_pasado
                      ,
               data=sevdb_2x,
               overall=FALSE, test=TRUE)

latex(s1, caption='Clinical data',
      exclude1=TRUE, npct='both', 
      digits=3,
      prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
      middle.bold=TRUE, long = TRUE,
      #legend.bottom = TRUE, #insert.bottom = TRUE, 
      what="%", html = TRUE
      ) #change here for LaTeX PDF
```

```{r, fig.width=6, fig.height=5, fig.align='center'}
#plot(s1, which='categorical')
#Key(0,0)
```

```{r, fig.width=8, fig.height=4, fig.align='center'}
#par(mfrow=c(1,2))
#plot(s1, which='continuous')
#par(mfrow=c(1,1))
```


## Ab profile subset

- **60 samples** from the Severe study dataset.

```{r}
subsev <- data_frame(
  codigo= c( "LIM2048", "LIM1025", "LIM1029", "LIM2035", "LIM1043", "LIM2017", "LIM1054", 
             "LIM2102", "LIM2010", "LIM1038", "LIM2050", "LIM2115", "LIM2043", "LIM1004", 
             "LIM2030", "LIM2071", "LIM1035", "LIM2104", "LIM1051", "LIM2107", "LIM1062", 
             "LIM2072", "LIM2074", "LIM1056", "LIM1050", "LIM2110", "LIM2026", "LIM1047", 
             "LIM1057", "LIM1016", "LIM1063", "LIM1044", "LIM2044", "LIM1037", "LIM2007", 
             "LIM2086", "LIM2042", "LIM2101", "LIM1045", "LIM2085", "LIM1048", "LIM2066", 
             "LIM2067", "LIM1021", "LIM2077", "LIM1061", "LIM2056", "LIM2034", "LIM2016", 
             "LIM2111", "LIM1030", "LIM2096", "LIM1064", "LIM2093", "LIM1041", "LIM2040", 
             "LIM1052", "LIM2013", "LIM1059", "LIM2089"),
  abnomic= TRUE
)

sevdb_3 <- dplyr::inner_join(sevdb_2,subsev, by="codigo")
#html(Hmisc::contents(sevdb_3), maxlevels=10, levelType='table')

sevdb_3 %>% group_by(malaria_pasado, malaria_pasado_num) %>% dplyr::count()
sevdb_3 %>% dplyr::count(malaria_pasado)
sevdb_3 %>% dplyr::count(sev_WHO)
#sevdb_3 %>% group_by(cas_con_admin2, sev_WHO) %>% dplyr::count()
sevdb_3 %>% group_by(malaria_pasado, sev_WHO) %>% dplyr::count()

#sevdb_3 %>% group_by(admin_uci_corr,dias_uci) %>% dplyr::count()
#sevdb_3 %>% group_by(sdra,lung_injury) %>%  dplyr::count()
#sevdb_3 %>% group_by(ictericia_corr,hiperbil) %>%  dplyr::count()
#sevdb_3 %>% group_by(anemia_severa, hemoglobinuria_corr) %>%  dplyr::count()
```

- Joint `parasitemia` from `28Nov2014_Ed.dta` **[WARNING]**

```{r, eval=FALSE}
sevdb_4 <- sevdb_3 %>% 
  dplyr::inner_join(sevdb_x, by="codigo")
#html(Hmisc::contents(sevdb_4), maxlevels=10, levelType='table')
```
```{r, eval=FALSE}
d <- describe(sevdb_4)
html(d, size=90, scroll=TRUE)
```


```{r, eval=FALSE}
# THIS IS IN eval=FALSE !!!!!!!
sevdb_4 <- sevdb_3 %>% 
  haven::as_factor() %>%
  mutate_at(vars(-c(1:7),-ends_with("_num"),-abnomic), as.factor) %>% 
  mutate(admin_uci_corr= fct_recode(admin_uci_corr,NULL="No sabe"),
         ictericia_corr= fct_recode(ictericia_corr,NULL="No sabe"),
         cns_malaria= fct_recode(cns_malaria,"Si"="1", "No"="0"),
         glas_11= fct_recode(glas_11,"Si"="1", "No"="0"),
         sdra= fct_recode(sdra,"Si"="1", "No"="0"),
         lung_injury= fct_recode(lung_injury,"Si"="1", "No"="0"),
         hiperbil= fct_recode(hiperbil,"Si"="1", "No"="0"),
         shock= fct_recode(shock,"Si"="1", "No"="0"),
         bleeding= fct_recode(bleeding,"Si"="1", "No"="0"),
         glic_normal= fct_recode(glic_normal,"Si"="1", "No"="0"),
         hiperparasitemia= fct_recode(hiperparasitemia,"Si"="1", "No"="0"),
         anemia_severa= fct_recode(anemia_severa,"Si"="1", "No"="0"),
         presenta_comorb= fct_recode(presenta_comorb,"Si"="1", "No"="0")) %>% 
  dplyr::inner_join(sevdb_x)

html(Hmisc::contents(sevdb_4), maxlevels=10, levelType='table')
```

### combinations of criterias
```{r}
# SUM SEVERITY CRITERIAS
sevdb_o <- sevdb_3 %>% 
  dplyr::rename(glas=glas_11,
         #sdra,
         lung=lung_injury,
         icter=ictericia_corr,
         bili=hiperbil,
         #shock,
         bleed=bleeding,
         glic=glic_normal,
         paras=hiperparasitemia,
         anemia=anemia_severa
         ) 
sevdb_o %>% 
  group_by(#cns_malaria,
           glas,
           sdra,
           lung,
           icter,
           bili,
           shock,
           bleed,
           glic,
           paras,
           anemia
           ) %>% 
  dplyr::count() %>% 
  dplyr::arrange(desc(n))
```
```{r}
sevdb_o %>% 
  select(#cns_malaria,
           glas,
           sdra,
           lung,
           icter,
           bili,
           shock,
           bleed,
           glic,
           paras,
           anemia
           ) %>% 
  summarise_all(sum,na.rm=T) #%>% t()
```

```{r, fig.height=4, fig.width=6, fig.align='center', eval=FALSE}
## symptoms x exposure

#plot(sevdb_3$malaria_pasado_num,sevdb_3$sev_WHO_num)
sevdb_3 %>% 
  group_by(malaria_pasado_num, sev_WHO_num) %>%
  dplyr::summarise(n=n()) %>%
  mutate(prop= n/sum(n)) %>%
  ggplot(aes(malaria_pasado_num, sev_WHO_num, colour=n)) + 
  geom_point() + 
  scale_color_viridis() +
  labs(title="sevdb_3")
```

```{r, eval=FALSE}
# **On severity**
s1 <- Hmisc::summaryM(edad + sexo +
                        malaria_pasado + #malaria_pasado_num + 
                        #sev_WHO_num #+ sev_WHO
                        fiebre_hoy_corr + presenta_comorb +
                        parasitemia #+
                      ~ sev_WHO #+ 
                      #malaria_pasado
                      ,
               data=sevdb_4,
               overall=FALSE, test=FALSE)

latex(s1, caption='Clinical data',
      exclude1=TRUE, npct='both', 
      digits=3,
      prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
      middle.bold=TRUE, long = TRUE,
      #legend.bottom = TRUE, #insert.bottom = TRUE, 
      what="%", html = TRUE
      ) #change here for LaTeX PDF
```

```{r, eval=FALSE}
s1 <- Hmisc::summaryM(glas_11 +
                        sdra +
                        lung_injury +
                        ictericia_corr + hiperbil +
                        shock +
                        bleeding +
                        glic_normal +
                        anemia_severa +
                        hiperparasitemia + #parasitemia + 
                        sev_WHO_num +
                        fiebre_hoy_corr
                      ~ sev_WHO
                      ,
               data=sevdb_4,
               overall=FALSE, test=FALSE)

latex(s1, caption='Laboratory data',
      exclude1=TRUE, npct='both', 
      digits=3,
      prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
      middle.bold=TRUE, long = TRUE,
      #legend.bottom = TRUE, #insert.bottom = TRUE, 
      what="%", html = TRUE
      ) #change here for LaTeX PDF
```

```{r, eval=FALSE}
# **On previous episodes of malaria**
s1 <- Hmisc::summaryM(edad + sexo + 
                        #malaria_pasado_num + #malaria_pasado +
                        sev_WHO + sev_WHO_num +
                        fiebre_hoy_corr + presenta_comorb +
                        parasitemia #+
                      ~ #sev_WHO #+ 
                      malaria_pasado
                      ,
               data=sevdb_4,
               overall=FALSE, test=FALSE)

latex(s1, caption='Clinical data',
      exclude1=TRUE, npct='both', 
      digits=3,
      prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
      middle.bold=TRUE, long = TRUE,
      #legend.bottom = TRUE, #insert.bottom = TRUE, 
      what="%", html = TRUE
      ) #change here for LaTeX PDF
```
```{r, eval=FALSE}
s1 <- Hmisc::summaryM(glas_11 +
                        sdra +
                        lung_injury +
                        ictericia_corr + hiperbil +
                        shock +
                        bleeding +
                        glic_normal +
                        anemia_severa +
                        hiperparasitemia + #parasitemia + 
                        sev_WHO_num +
                        fiebre_hoy_corr
                      ~ #sev_WHO #+ 
                      malaria_pasado
                      ,
               data=sevdb_4,
               overall=FALSE, test=FALSE)

latex(s1, caption='Laboratory data',
      exclude1=TRUE, npct='both', 
      digits=3,
      prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
      middle.bold=TRUE, long = TRUE,
      #legend.bottom = TRUE, #insert.bottom = TRUE, 
      what="%", html = TRUE
      ) #change here for LaTeX PDF
```

```{r, eval=FALSE}
#**On severity by previous episodes of malaria**
s1 <- Hmisc::summaryM(edad + sexo +
                        #malaria_pasado_num + #malaria_pasado +
                        #sev_WHO_num #+ sev_WHO
                        fiebre_hoy_corr + presenta_comorb +
                        parasitemia #+
                      ~ sev_WHO + 
                      malaria_pasado
                      ,
               data=sevdb_4,
               overall=FALSE, test=FALSE)

latex(s1, caption='Clinical data',
      exclude1=TRUE, npct='both', 
      digits=3,
      prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
      middle.bold=TRUE, long = TRUE,
      #legend.bottom = TRUE, #insert.bottom = TRUE, 
      what="%", html = TRUE
      ) #change here for LaTeX PDF
```

```{r, eval=FALSE}
s1 <- Hmisc::summaryM(glas_11 +
                        sdra +
                        lung_injury +
                        ictericia_corr + hiperbil +
                        shock +
                        bleeding +
                        glic_normal +
                        anemia_severa +
                        hiperparasitemia + #parasitemia + 
                        sev_WHO_num +
                        fiebre_hoy_corr
                      ~ sev_WHO + 
                      malaria_pasado
                      ,
               data=sevdb_4,
               overall=FALSE, test=FALSE)

latex(s1, caption='Laboratory data',
      exclude1=TRUE, npct='both', 
      digits=3,
      prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
      middle.bold=TRUE, long = TRUE,
      #legend.bottom = TRUE, #insert.bottom = TRUE, 
      what="%", html = TRUE
      ) #change here for LaTeX PDF
```

```{r, fig.width=7, fig.height=6, fig.align='center'}
#plot(s1, which='categorical')
#Key(0,0)
```

```{r, fig.width=8, fig.height=4, fig.align='center'}
#par(mfrow=c(1,2))
#plot(s1, which='continuous')
#par(mfrow=c(1,1))
```

```{r, eval=FALSE}
# appendix
bpplt()
```

# References