---
title: "Severe Covariates (w/ SEVERE_28nov2014.dta)"
author: "Andree Valle Campos"
date: '`r format(Sys.Date(), format="%d %B %Y")`'
output:
  html_notebook:
    toc: yes
    toc_float:
      collapsed: yes
  html_document:
    fig_caption: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
bibliography: malaria.bib
csl: american-medical-association.csl
---

<!-- last modification: 09set2016 -->

# Introduction
This report is based on the set of reproducible examples of the R[@R] `Hmisc` package[@Hmisc], put together in an `rmarkdown`[@rmarkdown] notebook using `RStudio`[@rstudio] and `knitr`[@knitr]. Click [here](http://data.vanderbilt.edu/fh/R/Hmisc/examples.nb.html) for more information.

# Setup 
```{r setup,results='hide', message=FALSE}
require(Hmisc)
require(plotly)
options(#grType='plotly',  # for certain graphics functions
        width = 110) # to expand the limits of CONSOLE output
mu <- markupSpecs$html   # markupSpecs is in Hmisc

# The following hidden command (<code>r mu$widescreen()</code>), causes the html notebook to use an entire wide screen.
#mu$widescreen()
```
`r mu$widescreen()`

## Severe dataset input
```{r sevdata, echo=FALSE}
# Extended
sevData<-read.table("SEVERA_covariates_224samples_EXTENDED_18ago2016.csv",row.names=1,header=TRUE, sep=",", 
                        colClasses = c(rep("factor",5),"numeric","factor","numeric","numeric",rep("factor",5),"numeric","numeric",
                                       rep("factor",5),rep("numeric",9),"factor","numeric",rep("factor",2),rep("numeric",2)))
# Extended
##
#summary(sevData)
sampleSEV <- sevData[c( "LIM2048", "LIM1025", "LIM1029", "LIM2035", "LIM1043", "LIM2017", "LIM1054", "LIM2102", "LIM2010", "LIM1038", "LIM2050", "LIM2115", "LIM2043", "LIM1004", "LIM2030", "LIM2071", "LIM1035", "LIM2104", "LIM1051", "LIM2107", "LIM1062", "LIM2072", "LIM2074", "LIM1056", "LIM1050", "LIM2110", "LIM2026", "LIM1047", "LIM1057", "LIM1016", "LIM1063", "LIM1044", "LIM2044", "LIM1037", "LIM2007", "LIM2086", "LIM2042", "LIM2101", "LIM1045", "LIM2085", "LIM1048", "LIM2066", "LIM2067", "LIM1021", "LIM2077", "LIM1061", "LIM2056", "LIM2034", "LIM2016", "LIM2111", "LIM1030", "LIM2096", "LIM1064", "LIM2093", "LIM1041", "LIM2040", "LIM1052", "LIM2013", "LIM1059", "LIM2089"),]
#summary(sampleSEV)
#head(sampleSEV)
#write.csv(sampleSEV, "samples_SEVERE_60_covariates_35.csv")
```


```{r sevdataLOGIC, echo=FALSE}
# ONE WAY
#sevFunction <- function(casecontrol) {if (casecontrol == "1") "Severe" else "Non-severe"}
#sampleSEV$casecontrol <- as.factor(unlist(lapply(sampleSEV$casecontrol, sevFunction)))
# SIMPLIFIED
levels(sampleSEV$casecontrol) <- c("Non-severe","Severe") # EVIDENCE: all pregnats were 2
levels(sampleSEV$sexo) <- c("hombre","mujer") # EVIDENCE: all pregnats were 2
#levels(sampleSEV$sin_enf_cronica) <- c("1","0") # BECAUSE: here 0 means presence of cronic disease!!! ¿CON O SIN?
#

# NEW CREATED variables --> WHO CRITERIA
sampleSEV$jaundice = factor(sampleSEV$bilitot > 2.5, labels= c("0", "1")) # 3 in Quispe, 2014 (paper), 2.5 in Smith(poster)
sampleSEV$renal = factor(sampleSEV$creatinina > 3, labels= c("0", "1"))
sampleSEV$sevAnemia = factor(sampleSEV$hemoglobina < 7, labels= c("0", "1")) # 5 if Severe Anemia
sampleSEV$hipoglicemia = factor(sampleSEV$glucosa < 40, labels= c("0"))
sampleSEV$cerebralX = factor(sampleSEV$glasgow < 11, labels= c("0", "1"))
#



#
# to avoid 98/99 maintaining "NA"
PREfunction <- function(covariate) {
  if (is.na(covariate) == TRUE) {
    NA
  } else if (covariate == "1") {
      "1"
  } else {
      "0"
    }}
#
for (i in c(1,3:5,7:ncol(sampleSEV))) {
  if (class(sampleSEV[,i]) == "factor") {
    sampleSEV[,i] <- as.factor(unlist(lapply(sampleSEV[,i], PREfunction)))
    #sampleSEV[,i] <- factor(unlist(lapply(sampleSEV[,i], PREfunction)), exclude = NULL)
    }
}
#

#
levels(sampleSEV$sdra) <- c("1", "0")
  sampleSEV$sdra <- relevel(sampleSEV$sdra, ref = "0")
levels(sampleSEV$postracion) <- c("1", "0")
  sampleSEV$postracion <- relevel(sampleSEV$postracion, ref = "0")
levels(sampleSEV$trombocitopenia) <- c("1", "0")
  sampleSEV$trombocitopenia <- relevel(sampleSEV$trombocitopenia, ref = "0")
levels(sampleSEV$hipoglicemia) <- c("0", "1")
#

# WHY THIS METHOD DOES NOT WORK?
#dfr7df <- subset(sampleSEV, select = c("jaundice", "renal", "sevAnemia", "hipoglicemia", "sdra", "shock", "cerebralX"))
#sampleSEV$CumSumSevere7 = as.numeric(apply(dfr7, 1, sum, na.rm=TRUE))
dfr7 <- cbind(sampleSEV$jaundice, sampleSEV$renal, sampleSEV$sevAnemia, sampleSEV$hipoglicemia, 
              sampleSEV$sdra, sampleSEV$shock, sampleSEV$cerebralX) -1
colnames(dfr7) <- c("jaundice", "renal", "sevAnemia", "hipoglicemia", "sdra", "shock", "cerebralX")
#dfr7 <- as.data.frame(dfr7)
#dfr7df == dfr7 #TRUE
sampleSEV$CumSumSevere7 = as.numeric(apply(dfr7, 1, sum, na.rm=TRUE))
#subset(sampleSEV, select = c("jaundice", "renal", "sevAnemia", "hipoglicemia", "sdra", "shock", "cerebralX", "CumSumSevere7"))

#
dfr10 <- cbind(sampleSEV$jaundice, sampleSEV$renal, sampleSEV$sevAnemia, sampleSEV$hipoglicemia, 
               sampleSEV$sdra, sampleSEV$shock, sampleSEV$cerebralX, 
               sampleSEV$postracion, sampleSEV$coma, sampleSEV$convulsion) -1 #
colnames(dfr10) <- c("jaundice", "renal", "sevAnemia", "hipoglicemia", "sdra", "shock", "cerebralX", 
                     "postracion", "coma", "convulsion") #

sampleSEV$CumSumSevere10 = as.numeric(apply(dfr10, 1, sum, na.rm=TRUE))
#

sampleSEVx <- sampleSEV
sampleSEVx$sample <- rownames(sampleSEVx)


#
# to CHANGE 1/0 to presence/absence
ONEfunction <- function(covariate) {
  if (is.na(covariate) == TRUE) {
    NA
  } else if (covariate == "1") {
      "present"
  } else {
      "absent"
    }}

for (i in c(1,3:5,7:ncol(sampleSEV))) {
  if (class(sampleSEV[,i]) == "factor") {
    sampleSEV[,i] <- as.factor(unlist(lapply(sampleSEV[,i], ONEfunction)))
    }
}
#

levels(sampleSEV$sdra) <- c("present", "absent")
  sampleSEV$sdra <- relevel(sampleSEV$sdra, ref = "absent")
levels(sampleSEV$postracion) <- c("present", "absent")
  sampleSEV$postracion <- relevel(sampleSEV$postracion, ref = "absent")
levels(sampleSEV$trombocitopenia) <- c("present", "absent")
  sampleSEV$trombocitopenia <- relevel(sampleSEV$trombocitopenia, ref = "absent")
levels(sampleSEV$hipoglicemia) <- c("absent", "present")

#
# TO EVALUATE
#table(sampleSEV$sdra, useNA = "ifany") #like trombocitopenia
#table(sampleSEV$postracion, useNA = "ifany")
#table(sampleSEV$convulsion, useNA = "ifany") #like coma
#table(sampleSEV$hipoglicemia, useNA = "ifany")
#eset$Group <- relevel(eset$Group, ref = "0")
#
# rm(list = ls())


####### On NUMBER of previous malaria events
#sampleSEV["LIM2017",]
#sampleSEV["LIM2017","malaria_pasado_num"] # 98? -> NA
sampleSEV["LIM2017","malaria_pasado_num"] <- NA      ## deliberate modification. [coincidently, also in "PVX_111175.544", because extremely low value]
#table(sampleSEV$malaria_pasado_num)
#summary(sampleSEV$malaria_pasado_num)
sampleSEV$con_exp = factor(sampleSEV$malaria_pasado_num >= 1, labels= c("Sin exposicion", "Con exposicion"))
#summary(sampleSEV$con_exp)


#summary(sampleSEV)
```


```{r internalTEST, eval=FALSE, echo=FALSE}
#
options(digits=3)
set.seed(3)
n <- 20
sex <- factor(sample(c("m","f"), n, rep=TRUE))
age <- rnorm(n, 50, 5)
treatment <- factor(sample(c("Drug","Placebo"), n, rep=TRUE))
# Generate a 3-choice variable; each of 3 variables has 5 possible levels
symp <- c('Headache','Stomach Ache','Hangnail',
          'Muscle Ache','Depressed')
symptom1 <- sample(symp, n, TRUE)
symptom2 <- sample(symp, n, TRUE)
symptom3 <- sample(symp, n, TRUE)
cbind(symptom1, symptom2, symptom3)[1:5,]
Symptoms <- mChoice(symptom1, symptom2, symptom3, label='Primary Symptoms')
Symptoms
print(Symptoms, long=TRUE)
format(Symptoms[1:5])
inmChoice(Symptoms,'Headache')
levels(Symptoms)
inmChoice(Symptoms, 3)
inmChoice(Symptoms, c('Headache','Hangnail'))
# Note: In this example, some subjects have the same symptom checked
# multiple times; in practice these redundant selections would be NAs
# mChoice will ignore these redundant selections

meanage <- N <- numeric(5)
for(j in 1:5) {
 meanage[j] <- mean(age[inmChoice(Symptoms,j)])
 N[j] <- sum(inmChoice(Symptoms,j))
}
names(meanage) <- names(N) <- levels(Symptoms)
meanage
N

# Manually compute mean age for 2 symptoms
mean(age[symptom1=='Headache' | symptom2=='Headache' | symptom3=='Headache'])
mean(age[symptom1=='Hangnail' | symptom2=='Hangnail' | symptom3=='Hangnail'])

summary(Symptoms)

#Frequency table sex*treatment, sex*Symptoms
summary(sex ~ treatment + Symptoms, fun=table)
# Check:
ma <- inmChoice(Symptoms, 'Muscle Ache')
table(sex[ma])

# could also do:
# summary(sex ~ treatment + mChoice(symptom1,symptom2,symptom3), fun=table)

#Compute mean age, separately by 3 variables
summary(age ~ sex + treatment + Symptoms)


summary(age ~ sex + treatment + Symptoms, method="cross")

f <- summary(treatment ~ age + sex + Symptoms, method="reverse", test=TRUE)
f
# trio of numbers represent 25th, 50th, 75th percentile
print(f, long=TRUE)
#
```

```{r mChoice1, echo=FALSE, eval=TRUE}

symptom10 <- subset(sampleSEVx, 
                  select = c("sdra", "shock", "cerebralX", "renal", "sevAnemia", 
                             "jaundice", "hipoglicemia", 
                             "postracion", "coma", "convulsion"))

colnames(symptom10) <- c("Lesion pulmonar", "Shock circulatorio","Malaria cerebral*", "Insuficiencia renal*", "Anemia Severa*", "Ictericia*",  "Hipoglicemia*", "Postracion", "Coma", "Convulsion")

#colnames(symptom10) <- c("Lesion pulmonar (SDRA)", "Shock circulatorio (PA sistolica < 80mmHg)","Malaria cerebral* (Glasgow < 11/15)", "Insuficiencia renal* (creatinina > 3mg/dL)", "Anemia Severa* (hemoglobina < 7g/dL)", "Ictericia* (bilirrubina > 2.5mg/dL)",  "Hipoglicemia* (glucosa < 40mg/dL)", "Postracion", "Coma", "Convulsion")


for (i in 1:10) {
  levels(symptom10[,i]) <- c("",colnames(symptom10)[i])
  symptom10[,i] <- as.character(symptom10[,i])
  symptom10[,i][is.na(symptom10[,i])] <- ""
}


#
CriticalSymp5 <- mChoice(symptom10$`Lesion pulmonar`, symptom10$`Shock circulatorio`, symptom10$`Malaria cerebral*`, symptom10$`Insuficiencia renal*`, symptom10$`Anemia Severa*`, label='Critical illness criteria')
#CriticalSymp5 <- mChoice(symptom10$`Lesion pulmonar (SDRA)`, symptom10$`Shock circulatorio (PA sistolica < 80mmHg)`, symptom10$`Malaria cerebral* (Glasgow < 11/15)`, symptom10$`Insuficiencia renal* (creatinina > 3mg/dL)`, symptom10$`Anemia Severa* (hemoglobina < 7g/dL)`, label='Critical illness criteria')
#summary(CriticalSymp5)

CriticalSymp2 <- mChoice(symptom10$`Ictericia*`, symptom10$`Hipoglicemia*`, label='Non-critical illness criteria')
#CriticalSymp2 <- mChoice(symptom10$`Ictericia* (bilirrubina > 2.5mg/dL)`, symptom10$`Hipoglicemia* (glucosa < 40mg/dL)`, label='Non-critical illness criteria')
#summary(CriticalSymp2)

CriticalSymp3 <- mChoice(symptom10$Postracion, symptom10$Coma, symptom10$Convulsion, label='Central Nervous System')
#summary(CriticalSymp3)

CriticalSymp7 <- mChoice(symptom10$`Lesion pulmonar`, symptom10$`Shock circulatorio`, symptom10$`Malaria cerebral*`, symptom10$`Insuficiencia renal*`, symptom10$`Anemia Severa*`, symptom10$`Ictericia*`, symptom10$`Hipoglicemia*`, label='WHO Severe criteria 7')


CriticalSymp10 <- mChoice(symptom10$`Lesion pulmonar`, symptom10$`Shock circulatorio`, symptom10$`Malaria cerebral*`, symptom10$`Insuficiencia renal*`, symptom10$`Anemia Severa*`, symptom10$`Ictericia*`, symptom10$`Hipoglicemia*`, symptom10$Postracion, symptom10$Coma, symptom10$Convulsion, label='WHO Severe criteria 10')


```


# Data Dictionary
`upData` is used to

- create a new variable from an old one
- add labels to variables
- add units to variables
- remove the old variables
- automatically move units of measurements from parenthetical expressions in labels to separate `units` attributed used by `Hmisc` and `rms` functions for table making and graphics

`contents` is used to print a data dictionary, run through an `html` method for nicer output.
```{r sevUPdata, results='asis', echo=FALSE}
#
#AGEfunction <- function(x) {
#  if (x < 18) {
#    "< 18"
#  } else if (x >= 65) {
#      ">= 65"
#  } else {
#      "18-64"
#  }}#
AGEfunction <- function(x) {cut(x, c(0,15,25,35,45,60,100))}
#
# Have upData move units from labels to separate attribute
pbc <- upData(sampleSEV,
              cont_plaquetas = cont_plaquetas / 1000,
              edad_CAT = AGEfunction(edad),
              
              labels = c(edad = "Edad", edad_CAT = "Edad*", sexo = "Sexo", sin_enf_cronica = "Sin enfermedad cronica", 
                         malaria_pasado_num = "Eventos previos de malaria", 
                         con_exp = "Exposicion previa a malaria", casecontrol = "Sintomatologia",
                         
                         hemoglobina = "Hemoglobina", creatinina= "Creatinina", glucosa= "Glucosa",
                         bilitot= "Bilirrubina total", bilind= "Bilirrubuna indirecta", 
                         bildirect= "Bilirrubina directa",
                         parasitemia= "Parasitemia circulante", eritro_parasit = "eritrocitos parasitados",
                         glasgow = "Glasgow", cont_plaquetas = "Conteo de plaquetas", 
                         
                         postracion = "-postracion", coma= "-coma", convulsion = "-convulsion",
                         
                         admin_uci = "Admision a UCI", jaundice = "6.Ictericia*", 
                         shock= "2.Shock circulatorio",
                         cerebralX = "3.Malaria cerebral?*", renal = "4.Insuficiencia renal*", 
                         sevAnemia = "5.Anemia Severa*",
                         sdra= "1.Lesion pulmonar", hipoglicemia = "7.Hipoglicemia*",
                         
                         tx_pasado = "¿tx_pasado?", criticalnopros = "[¿criticalnopros?]",
                         criticos = "[criticos]", cerebral2 = "[cerebral2]",
                         anemia_severa = "¿anemia severa?", trombocitopenia = "¿trombocitopenia?",
                         
                         CumSumSevere7 = "Pacientes con #/7 criterios", 
                         CumSumSevere10 = "Pacientes con #/10 criterios"),
              
              units = c(edad = "años", edad_CAT = "años",
                        hemoglobina = "g/dL", creatinina= "mg/dL", glucosa= "mg/dL", 
                        bilitot= "mg/dL", bilind= "mg/dL", bildirect= "mg/dL",
                        parasitemia= "par/uL", cont_plaquetas = "x1000/uL", 
                        eritro_parasit = "¿%?",
                        
                        shock = "PA sistolica < 80mmHg", sdra= "SDRA", cerebralX = "Glasgow < 11/15", 
                        jaundice = "bilirrubina > 2.5mg/dL", renal = "creatinina > 3mg/dL",
                        sevAnemia = "hemoglobina < 7g/dL", hipoglicemia = "glucosa < 40mg/dL",
                        
                        criticalnopros = "[¿?]", criticos = "[¿?]", cerebral2 = "[*]",
                        anemia_severa = "[¿?]", trombocitopenia = "[¿?]"),
              #moveUnits=TRUE, 
              html=TRUE)
#
html(contents(pbc), maxlevels=10, levelType='table')
```

```{r sevUPdataLATEX, results='asis', echo=FALSE, eval=FALSE}
#
#AGEfunction <- function(x) {
#  if (x < 18) {
#    "< 18"
#  } else if (x >= 65) {
#      ">= 65"
#  } else {
#      "18-64"
#  }}#
AGEfunction <- function(x) {cut(x, c(0,15,25,35,45,60,100))}
#
# Have upData move units from labels to separate attribute
pbcLATEX <- upData(sampleSEV,
              cont_plaquetas = cont_plaquetas / 1000,
              edad_CAT = AGEfunction(edad),
              
              labels = c(cont_plaquetas = "Conteo de plaquetas (x1000/uL)",
                         hemoglobina = "Hemoglobina (g/dL)",
                         creatinina= "Creatinina (mg/dL)",
                         glucosa= "Glucosa (mg/dL)",
                         bilitot= "Bilirrubina total (mg/dL)",
                         bilind= "Bilirrubuna indirecta (mg/dL)",
                         bildirect= "Bilirrubina directa (mg/dL)",
                         parasitemia= "Parasitemia circulante (par/uL)",
                         
                         postracion = "-postracion", 
                         coma= "-coma", 
                         convulsion = "-convulsion",
                         
                         shock= "2.Shock (PA sistolica < 80mmHg)",
                         sdra= "1.Lesion pulmonar (SDRA)",
                         cerebralX = "3.Malaria cerebral?* (Glasgow < 11/15)",
                         glasgow = "Glasgow",
                         
                         tx_pasado = "¿tx_pasado?",
                         criticalnopros = "[¿criticalnopros?]",
                         criticos = "[criticos]",
                         cerebral2 = "[cerebral2]",
                         
                         anemia_severa = "¿anemia severa?",
                         trombocitopenia = "¿trombocitopenia?",
                         
                         admin_uci = "Admision a UCI",
                         sin_enf_cronica = "Sin enfermedad cronica",
                         malaria_pasado_num = "Eventos previos de malaria", 
                         con_exp = "Exposicion previa a malaria", casecontrol = "Sintomatologia",
                         eritro_parasit = "eritrocitos parasitados (%?)",
                         
                         edad = "Edad", edad_CAT = "Edad*",
                         sexo = "Sexo",
                         
                         jaundice = "6.Ictericia* (bilirrubina > 2.5mg/dL)",
                         renal = "4.Insuficiencia renal aguda* (creatinina > 3mg/dL)",
                         sevAnemia = "5.Anemia Severa* (hemoglobina < 7g/dL)",
                         hipoglicemia = "7.Hipoglicemia* (glucosa < 40mg/dL)",
                         CumSumSevere7 = "Pacientes con #/7 criterios",
                         CumSumSevere10 = "Pacientes con #/10 criterios"),
              #moveUnits=TRUE, 
              html=TRUE)
##

```


# Descriptive Statistics Without Stratification
The html method is used for the `describe` function, and the output is put in a scrollable box.  But the graphical display of the descriptives statistics that follows this is probably better.

```{r sevdescribe, fig.height=8}
d <- describe(pbc)
html(d, size=80, scroll=TRUE)
p <- plot(d)
prList(p)
```


```{r plotlym, fig.height=8}
htmltools::tagList(lapply(p, plotly::as.widget))
```


# Stratified Descriptive Statistics
Stratified quantiles, means/SD, and proportions per group are produced. Results are plotted before rendering as an advanced html table:

- categorical variables: a single dot chart
- continuous variables: a series of extended box plots

As before, the html method is used to `describe` each group (Severe and Non-severe).
```{r sevTREATMENTdescribe}
s <- describe(subset(pbc, casecontrol == "Severe"))
html(s, size=80, scroll=TRUE)

ns <- describe(subset(pbc, casecontrol != "Severe"))
html(ns, size=80, scroll=TRUE)
# EN sampleSEV$malaria_pasado_num --> paciente de 72 años c/ 98 ¿?
```



## CLINICAL data
```{r sev2summaryMa, warning=FALSE, fig.height=4}
s2 <- summaryM(edad + edad_CAT + sexo + sin_enf_cronica + con_exp + malaria_pasado_num #+ tx_pasado + # general
                #admin_uci + postracion + coma + convulsion + 
                 #shock + sdra + glasgow + cerebralX # clinical
              ~ casecontrol,
              data=pbc,
              #pbc$drug!="not randomized",
							overall=FALSE, test=TRUE)
plot(s2, which='categorical')
Key(0,1)
```

```{r sev2summaryMb, warning=FALSE,  fig.width=8, fig.height=4}
par(mfrow=c(1,2))
plot(s2, which='continuous')
par(mfrow=c(1,1))
```

```{r sev2summaryM3, eval=TRUE}
latex(s2, caption='Clinical data',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, #insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF
```

```{r sev2summaryM3pdf, eval=FALSE, echo=FALSE}
s2 <- summaryM(edad + edad_CAT + sexo + sin_enf_cronica +  con_exp + malaria_pasado_num #+ tx_pasado + # general
                #admin_uci + postracion + coma + convulsion + 
                 #shock + sdra + glasgow + cerebralX # clinical
              ~ casecontrol,
              data=pbcLATEX,
              #pbc$drug!="not randomized",
							overall=FALSE, test=TRUE)
# alternative
y <- latex(s2, caption='Clinical data',
     exclude1=TRUE, npct='both', 
     digits=3,
     #prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
     middle.bold=TRUE, #long = TRUE,
     legend.bottom = TRUE, #insert.bottom = TRUE, 
     what="%", ctable=TRUE,
     html = FALSE) #change here for LaTeX PDF
dvi(y, width=7, height=9) # ANY OTHER ALTERNATIVE? Why not a HORIZONTAL page?
```

***
### **Previous exposure** to Malaria 
(Non-severe vs Severe)
```{r sev3summaryM3perWITHexpo, eval=TRUE, warning=FALSE, fig.height=4}

table(sampleSEV$malaria_pasado_num)
summary(sampleSEV$con_exp)

s3expo <- summaryM(edad + edad_CAT + sexo + sin_enf_cronica +  casecontrol + malaria_pasado_num 
                   ~ con_exp,
              data=pbc,
							overall=FALSE, test=TRUE)
plot(s3expo, which='categorical')
Key(0,0)
#par(mfrow=c(2,3))
#plot(s3age1, which='continuous')
#par(mfrow=c(1,1))
#
```

```{r sev3summaryM3perWITHexpob, warning=FALSE,  fig.width=8, fig.height=4}
par(mfrow=c(1,2))
plot(s3expo, which='continuous')
par(mfrow=c(1,1))
```

```{r sev3summaryM3perWITHexpoc, eval=TRUE}
latex(s3expo, caption='Clinical data (exposure)',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, #insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF
```


***

## LABORATORY data
```{r sev3summaryMa}
s3 <- summaryM(anemia_severa + hemoglobina + creatinina + # lab critical
                glucosa + bilitot + bilind + bildirect + # lab non-critical
                parasitemia + eritro_parasit + trombocitopenia + cont_plaquetas  # covariates
              ~ casecontrol,
              data=pbc,
							overall=FALSE, test=TRUE)
#plot(s3, which='categorical')
```

```{r sev3summaryMb, warning=FALSE,  fig.width=10, fig.height=10}
par(mfrow=c(3,3))
plot(s3, which='continuous')
par(mfrow=c(1,1))
```

```{r sev3summaryM3, eval=TRUE}
latex(s3, caption='Laboratory data',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2,
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF
```

***
### **Previous exposure**
```{r sev3summaryMaEXPO}
s3expoLAB <- summaryM(hemoglobina + creatinina + # lab critical
                glucosa + bilitot + bilind + bildirect + # lab non-critical
                parasitemia + eritro_parasit + cont_plaquetas  # covariates
              ~ con_exp,
              data=pbc,
							overall=FALSE, test=TRUE)
#plot(s3, which='categorical')
```

```{r sev3summaryMbEXPO, warning=FALSE,  fig.width=10, fig.height=10}
par(mfrow=c(3,3))
plot(s3expoLAB, which='continuous')
par(mfrow=c(1,1))
```

```{r sev3summaryM3expo, eval=TRUE}
latex(s3expoLAB, caption='Laboratory data (exposure)',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2,
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF
```


***
### Age distribution 
(Non-severe vs Severe)
```{r sev3summaryM3perAGEcat, eval=TRUE, warning=FALSE,  fig.width=10, fig.height=5}
s3age1 <- summaryM(hemoglobina + creatinina + glucosa
              ~ casecontrol + edad_CAT,
              data=pbc,
              groups = "edad_CAT",
							overall=FALSE, test=TRUE)
par(mfrow=c(2,3))
plot(s3age1, which='continuous')
#par(mfrow=c(1,1))
#
s3age2 <- summaryM(bilitot + bilind + bildirect
              ~ casecontrol + edad_CAT,
              data=pbc,
              groups = "edad_CAT",
							overall=FALSE, test=TRUE)
#par(mfrow=c(2,3))
plot(s3age2, which='continuous')
#par(mfrow=c(1,1))
#
s3age3 <- summaryM(parasitemia + eritro_parasit + cont_plaquetas  # covariates
              ~ casecontrol + edad_CAT,
              data=pbc,
              groups = "edad_CAT",
							overall=FALSE, test=TRUE)
#par(mfrow=c(2,3))
plot(s3age3, which='continuous')
#par(mfrow=c(1,1))
```

### < 18
```{r sev3summaryM3perAGEcatXyoung, eval=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=6}
s3ageY <- summaryM(hemoglobina + creatinina + # lab critical
                glucosa + bilitot + bilind + bildirect + # lab non-critical
                parasitemia + eritro_parasit + cont_plaquetas  # covariates
              ~ casecontrol,
              data=subset(pbc, edad_CAT=="< 18"),
							overall=FALSE, test=TRUE,
							continuous = 2)

plot(s3ageY, which='continuous')
```

### 18-64
```{r sev3summaryM3perAGEcatXadult, eval=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=6}
s3ageA <- summaryM(hemoglobina + creatinina + # lab critical
                glucosa + bilitot + bilind + bildirect + # lab non-critical
                parasitemia + eritro_parasit + cont_plaquetas  # covariates
              ~ casecontrol,
              data=subset(pbc, edad_CAT=="18-64"),
							overall=FALSE, test=TRUE)

plot(s3ageA, which='continuous')
```

### >= 65
```{r sev3summaryM3perAGEcatXold, eval=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=6}
s3ageO <- summaryM(hemoglobina + creatinina + # lab critical
                glucosa + bilitot + bilind + bildirect + # lab non-critical
                parasitemia + eritro_parasit + cont_plaquetas  # covariates
              ~ casecontrol,
              data=subset(pbc, edad_CAT==">= 65"),
							overall=FALSE, test=TRUE,
							continuous = 2)

plot(s3ageO, which='continuous')
```

***
```{r sev3summaryM3perAGEcatX, eval=FALSE, warning=FALSE, echo=FALSE}
s3age <- summaryM(hemoglobina + creatinina + # lab critical
                glucosa + bilitot + bilind + bildirect + # lab non-critical
                parasitemia + eritro_parasit + cont_plaquetas  # covariates
              ~ casecontrol + edad_CAT,
              data=pbc,
              groups = "edad_CAT",
							overall=FALSE, test=TRUE)

#plot(s3age, which='continuous')

latex(s3age, caption='Laboratory data',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2,
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF

latex(s3ageY, caption='Laboratory data (< 18)',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2,
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF

latex(s3ageA, caption='Laboratory data (18-64)',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2,
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF

latex(s3ageO, caption='Laboratory data (>= 65)',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2,
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF

```
***
### PDF
```{r sev3summaryM3pdf, eval=FALSE, echo=FALSE}
s3 <- summaryM(anemia_severa + hemoglobina + creatinina + # lab critical
                glucosa + bilitot + bilind + bildirect + # lab non-critical
                parasitemia + eritro_parasit + trombocitopenia + cont_plaquetas  # covariates
              ~ casecontrol,
              data=pbcLATEX,
							overall=FALSE, test=TRUE)
# alternative
y <- latex(s3, caption='Laboratory data',
     exclude1=TRUE, npct='both', 
     digits=3,
     #prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
     middle.bold=TRUE, long = TRUE,
     legend.bottom = TRUE, #insert.bottom = TRUE, 
     what="%", ctable=TRUE,
     html = FALSE) #change here for LaTeX PDF
print(y)
dvi(y, width=7, height=9) # ANY OTHER ALTERNATIVE? Why not a HORIZONTAL page?
```




## WHO criteria

According to the most recent WHO report[@WHOsevere2014], for epidemiological and research purposes, **Severe Malaria** is defined as one or more of the following, occurring in the absence of an identified alternative cause, and in the presence of asexual parasitaemia:

```{r sev1summaryMxx, warning=FALSE, fig.height=9, fig.width=10}
s1xx <- summaryM(admin_uci + con_exp + # CRITICAL CARE
                 CriticalSymp3 + CriticalSymp5 + CriticalSymp2 + parasitemia +
                 CumSumSevere7 + CumSumSevere10
              ~ casecontrol,
              data=pbc,
							overall=FALSE, test=TRUE)
plot(s1xx, which='categorical')
Key(0,0)
#legend("bottomleft", legend = c("Severe", "Non-severe"), col = c("white", "black"), pch = 19) #WHY NOT WORK?
#plot(s1, which='continuous')
```

```{r sev1summaryM3xx, warning=FALSE}
latex(s1xx, caption='WHO criteria for severe malaria: Clinical and Laboratory data',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2,
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF
```


### **Previous exposure**
```{r sev1summaryMexpo, warning=FALSE, fig.height=9, fig.width=10}
s1expo <- summaryM(admin_uci + casecontrol + # CRITICAL CARE
                 CriticalSymp3 + CriticalSymp5 + CriticalSymp2 + parasitemia +
                 CumSumSevere7 + CumSumSevere10
              ~ con_exp,
              data=pbc,
							overall=FALSE, test=TRUE)
plot(s1expo, which='categorical')
Key(0,0)
#plot(s1, which='continuous')
```

```{r sev1summaryM3expo, warning=FALSE}
latex(s1expo, caption='WHO criteria for severe malaria: Clinical and Laboratory data (exposure)',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2,
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF
```


***
### NA considered*
```{r sev1summaryM, warning=FALSE, fig.height=9, fig.width=10}
s1 <- summaryM(cerebral2 + criticalnopros + criticos + # clasifications
                 admin_uci + con_exp +# CRITICAL CARE
                 postracion + coma + convulsion + 
                 sdra + shock + cerebralX + renal + sevAnemia + 
                 jaundice + hipoglicemia + 
                 parasitemia + # NOT-PRACTICAL
                 CumSumSevere7 + CumSumSevere10
              ~ casecontrol,
              data=pbc,
							overall=FALSE, test=TRUE)
#plot(s1, which='categorical')
#plot(s1, which='continuous')
```

```{r sev1summaryM3, warning=FALSE}
latex(s1, caption='WHO criteria for severe malaria: Clinical and Laboratory data',
     exclude1=TRUE, npct='both', 
     digits=3,
     prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2,
     middle.bold=TRUE, long = TRUE,
     #legend.bottom = TRUE, insert.bottom = TRUE, 
     what="%", 
     html = TRUE) #change here for LaTeX PDF
```
***


```{r sev1summaryM3pdf, eval=FALSE, echo=FALSE}
s1 <- summaryM(#cerebral2 + criticalnopros + criticos + # clasifications
                 admin_uci + con_exp +# CRITICAL CARE
                 postracion + coma + convulsion + 
                 sdra + shock + cerebralX + renal + sevAnemia + 
                 jaundice + hipoglicemia +  # NON-CRITICAL ILLNESS (more colateral than life threatening)
                 #parasitemia + # NOT-PRACTICAL
                 CumSumSevere7 + CumSumSevere10
              ~ casecontrol,
              data=pbcLATEX,
							overall=FALSE, test=TRUE)
plot(s1, which='categorical')

# alternative -----------------------------------------------------------------------> NOW WORKING!
y <- latex(s1, caption='WHO criteria for severe malaria: Clinical and Laboratory data',
     exclude1=TRUE, npct='both', 
     digits=3,
     #prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
     middle.bold=TRUE, long = TRUE,
     legend.bottom = TRUE, #insert.bottom = TRUE, 
     what="%", ctable=TRUE,
     html = FALSE) #change here for LaTeX PDF
dvi(y, width=7, height=9)
```

```{r sev1summaryM3pdfII, warning=FALSE, eval=FALSE, echo=FALSE}
s1xx <- summaryM(admin_uci + con_exp +# CRITICAL CARE
                 CriticalSymp3 + CriticalSymp5 + CriticalSymp2 +
                 CumSumSevere7 + CumSumSevere10
              ~ casecontrol,
              data=pbc,
							overall=FALSE, test=TRUE)
plot(s1xx, which='categorical')

# alternative -----------------------------------------------------------------------> NOW WORKING!
y <- latex(s1xx, caption='WHO criteria for severe malaria: Clinical and Laboratory data',
     exclude1=TRUE, npct='both', 
     digits=3,
     #prmsd=TRUE, brmsd=TRUE, msdsize=mu$smaller2, #NOT-EVALUATE if PDF
     middle.bold=TRUE, long = TRUE,
     legend.bottom = TRUE, #insert.bottom = TRUE, 
     what="%", ctable=TRUE,
     html = FALSE) #change here for LaTeX PDF
dvi(y, width=7, height=9)

```



### Samples
```{r review, warning=FALSE}
# TO EVALUATE
ssubset <- subset(sampleSEVx, 
                  casecontrol== "Severe", 
                  select = c("sample",
                  "casecontrol","jaundice", "renal", "sevAnemia", "hipoglicemia", "sdra", 
                  "shock", "cerebralX", "CumSumSevere7", "postracion", "coma", "convulsion", "CumSumSevere10"))

nssubset <- subset(sampleSEVx,
                   casecontrol!= "Severe", 
                   select = c("sample",
                  "casecontrol","jaundice", "renal", "sevAnemia", "hipoglicemia", "sdra", 
                  "shock", "cerebralX", "CumSumSevere7", "postracion", "coma", "convulsion", "CumSumSevere10"))


ssubset
nssubset

```

### Symptoms
```{r symptomsCum}

CriticalSymp10
summary(CriticalSymp10)

CriticalSymp7
summary(CriticalSymp7)


CriticalSymp3
summary(CriticalSymp3)
CriticalSymp5
summary(CriticalSymp5)
CriticalSymp2
summary(CriticalSymp2)



```

## Useful variables

**Symptomatology, Previous experience and Age**

An important missing variable is *time since the last infection*[@Helb2015exposure] or *travel history* (time outside of the endemic area).But, since Iquitos is a *low transmission setting* (EIR?, PR?), lost of exposure to the infection could be more frequent even without living the endemic area (?).
```{r strates}
pbc$sample <- rownames(pbc)
sampleSEVcovariates <- pbc[,c("sample","casecontrol", "CumSumSevere7", "CumSumSevere10", "con_exp", "malaria_pasado_num", "edad_CAT", "edad")]
#sampleSEVcovariates
sampleSEVcovariates[order(sampleSEVcovariates$CumSumSevere10, decreasing = TRUE),]

#write.csv(sampleSEVcovariates, file = "sampleSEVcovariates.csv")

```


```{r extended}
bpplt()
```




# Computing Environment {#compenv}
`r mu$session()`




# References


<!---
To update html notebook on the server: cdatar examples.nb.html Hmisc
-->

